Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-06 manuscript/manuscript.tex
revisions to first section of Results

\subsection*{Substitution models}

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Comparison of substitution model features.}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

\skhcomment{Should this all be in the introduction?}

In nature, viral proteins evolve under both purifying selection, to maintain structure and function, and diversifying selection, to escape immune pressure. 
However, neither one of these constraints is homogenous across the entire protein. 
Some protein sites appear to be tolerant of many mutations, and sample many amino acids, while others appear to be mutationally tolerant, and sample only a select few. 
When modeling evolution it is important to account for this heterogeneity and phylogenetic substitution models do this in a variety of ways.

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution (\cite{yang1994maximum, yang2000codon}, \ref{fig:model_feature}B). 
Under this model, some sites would have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves and all non-synonymous changes at at given site are treated equally. 
In contrast, models from the mutation-selection family \cite{halpern1998evolutionary} account for purifying selection by explicitly defining the frequency of each amino acid at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A).

The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While most codon models have $<$ 20 parameters, mutation-selection models will have more than 10,000 parameters for an averaged size protein. 
As with all parameter-rich models, it is important to invoke some strategy to avoid overfitting when using a mutation-selection model.
We chose to define the amino-acid frequency parameters \textit{a priori} from a high-throughput assay called deep mutational scanning. 
The resulting model, an Experimentally Informed Codon Model (ExpCM) \citep{bloom2014experimentally, hilton2017phydms}, has a parameter space comparable to the other codon models.

These two strategies to account for site-to-site rate variation are not mutually exclusive. 
While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).

>>>>

\subsection*{Different ways hat substitution models account for purifying selection}

\jdbcomment{Some other comments on this section, which I think in general is pretty good: $\omega$ shows up in the figure, but is never mentioned in the text. I feel like GY94 models need to be explained at least in terms of what they and $\Gamma$ distribution stand for. Just introducing nomenclature. I feel like the last paragraph should tie back to panel C of the figure.}

\begin{figure}[H]
\centerline{\includegraphics[width=0.90\textwidth]{figures/model_feature.pdf}}
\caption{\label{fig:model_feature}
\textbf{Different ways of modeling site-to-site variation in purifying selection.}
\jdbcomment{Panels A and B need to be switched to match order in text. Then text updated too}
(A) Substitution model stationary states can either be identical at every site in the protein or allow to vary from site to site. 
(B) The relative rate of non-synonymous change, $\omega$, can be defined as one, gene-wide average or allowed to vary following some statistical distribution.
In order to achieve computational tractability, the distribution is discretized into $K$ bins and $\omega$ is set to the mean of each bin. 
A $\Gamma$ distribution with $K=4$ bins is shown here. 
(C) Substitution models can both, one, or neither of these features. 
We used models from the GY94 and the ExpCM families. 
}
\end{figure}

Proteins evolve under both purifying selection to maintain their structure and function. 
This purifying selection is not homogenous across sites in a protein.
It is also not homogenous across the different amino acids at a given site.
For instance, some protein sites strongly prefer hydrophobic amino acids, others may be constrained to just one or a few amino acids, and yet others may tolerate many amino acids.
In general, these constraints are highly idiosyncratic among sites, and so pose a challenge for phylogenetic substitution models.

The most common strategy to model rate variation is to allow the rate of non-synonymous change to vary among sites following some statistical distribution \citep[\ref{fig:model_feature}B;][]{yang1994maximum, yang2000codon}. 
Under such models, some sites have a higher rate of non-synonymous change than others. 
However, this rate is agnostic to the amino-acid identities themselves, and so all non-synonymous changes at at given site are treated equally.

In contrast, so-called ``mutation-selection'' models~\citep{halpern1998evolutionary} account for purifying selection by explicitly defining a different set of amino-acid preferences at each site in the protein. 
This more mechanistic formulation, without an arbitrary statistical distribution, results in a site-specific stationary state (\ref{fig:model_feature}A). 
These models better capture the site-to-site differences in amino-acid composition that is an obvious features of real proteins, and indeed they generally better describe actual evolution than models with only rate variation~\jdbcomment{cite Rodrigue / Lartillot papers, our papers, etc}.

The specificity of mutation-selection models comes at a cost in the form of an increased number of parameters. 
While codon substitution models with uniform stationary states typically have $<$20 parameters, mutation-selection models must specify 19 parameters for \emph{each} site (the stationary state is for 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, which leads to $\sim 10^4$ parameters for a typical size protein.
As with all parameter-rich models, it is important to obtain values for these parameters using some method that avoids overfitting.
Here we will primarily use experimentally informed codon models (ExpCM) which define values for these parameters \textit{a priori} from deep mutational scanning experiments so they do not need to be fit from phylogenetic data \jdbcomment{include citations}.
Therefore, for an ExpCM the number of remaining free parameters that are fit from the phylogenetic data are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian or maximum-likelihood estimation are discussed in the last section of the Results.

These two strategies to account for site-to-site variation in purifying selection are not mutually exclusive. 
\jdbcomment{I'm not sure if I love the rest of this paragraph. I feel like GY94 needs to be introduce more directly---perhaps in the paragraph beginning the ``The most common strategy...'' above. Also, it's a bit unclear what ``most appropriate'' means.}
While differences in stationary state manifest as differences in non-synonymous rates across sites, these models are most appropriate for modeling purifying selection \citep{spielman2015relationship}.
Proteins which are under strong purifying and diversifying selection may be better modeled by a combination of the two. 
For this study, we used four models from the GY94 and ExpCM families, which together represent all possible combinations of these two strategies (\ref{fig:model_feature}C).
